---
title: "Black bear hunting activity and harvests - Open Ontario"
execute:
    freeze: auto  # re-render only when source changes
    # do this to only re-render if the qmd file changes.
    # note that this is a little different from "cache: true"
    jupyter: python3
author: "Dean Jayatilleke, Junpu Xie, and Dave Campbell"
date: '2026-05-14'
slug: wild-life-data-balck-bear-hunting-activity-and-harvests-data-catalogue-from-goverrnment-of-ontario
categories: ["Regression", "Time Series",  "Open Ontario", "CKAN API","Software:Python","Software:R"]
subtitle: ''
summary: 'The dataset records numbers of black bears harvested and active black bear license holders every year in different wildlife management unit (WMU).'
output: html_document
ipynb-shell-interactivity: all
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



## Data Provider

The provincial government of Ontario provides open access to thousands of data sets via their [Open Data Ontario](https://data.ontario.ca/en/) portal. The purpose of sharing all data documents with the public is to remain transparent and accessible.  More details about the data license, training materials, and other information can be found [here](https://data.ontario.ca/about).

***

## Black Bear Hunting Activity

The dataset records numbers of black bears harvested and active black bear license holders every year from 2012 to 2018 in different wildlife management unit (WMU). The WMUs are the administrative coverage area that serves as a land base for wildlife monitoring and management.  

The original dataset of bears harvested and active hunters and its supporting document can be found [here](https://data.ontario.ca/dataset/black-bear-hunting-activity-and-harvests) and you can quickly preview the CSV dataset file [here](https://data.ontario.ca/dataset/black-bear-hunting-activity-and-harvests/resource/7dd6328e-74cc-4291-a041-2345cf7c6186). Moreover, The legend description can be viewed [here](https://files.ontario.ca/data-dictionary-black-bear-en.xlsx).


## Libraries

::: {.panel-tabset group="language"}

## Python
```{python plib}
import pandas as pd 
import requests
import seaborn as sb
import matplotlib as plt
```
## R
```{r rlib, message = FALSE, warning = FALSE}

library(jsonlite)
library(latticeExtra)
library(tidyverse) 
```

:::


### Finding the DataStore Resource Id

The Ontario Data Catalogue uses the [CKAN API](https://data.ontario.ca/about#developers). A CKAN package is a dataset landing page, while a resource is an individual file, table, API-backed data object, or supporting document attached to that dataset.

Before downloading the records, we can use the API to search for the black bear hunting dataset and identify the active CSV resource id needed by the `datastore_search` endpoint.  Here we'll search for **black bear** and see if we can find the data table identifier.   Search is performed by adding search terms to the `api_base` to construct a url.  We include a `timeout` in the request to avoid waiting forever if our internet (or the API) breaks.  The API returns a json file `search_results` with 10 rows (user defined value) that needs reshaping to make it easier to read the important values.

::: {.panel-tabset group="language"}

## Python
```{python psearch}

api_base = "https://data.ontario.ca/en/api/3/action"
search_terms = "black bear"

search_response = requests.get(
    f"{api_base}/package_search",
    params={
        "q": search_terms,
        "rows": 10
    },
    timeout=30
)
search_response.raise_for_status()

search_results = search_response.json()["result"]["results"]

datasets = pd.DataFrame([
    {
        "name": dataset["name"],
        "title": dataset["title"],
        "resources": dataset["num_resources"],
        "modified": dataset["metadata_modified"]
    }
    for dataset in search_results
])

datasets

```

## R
```{r rsearch, message = FALSE, warning = FALSE}

api_base <- "https://data.ontario.ca/en/api/3/action"
search_terms <- "black bear"

search_url <- paste0(
  api_base,
  "/package_search?q=", URLencode(search_terms, reserved = TRUE),
  "&rows=10"
)

search_results <- fromJSON(search_url, simplifyDataFrame = TRUE)$result$results

datasets <- search_results |>
  transmute(
    name,
    title,
    resources = num_resources,
    modified = metadata_modified
  )

datasets

```
:::

The search results include the dataset name `black-bear-hunting-activity-and-harvests`. That name can be passed as the value of `package_show` in the next API call.  The return is again in json format with the full metadata for the dataset, including last modification date, data license information, notes (here in both English and French), keywords all attached resources.  Our goal is to make sure that we obtain the right version of the data by finding the `resource_id` for the table. 

::: {.panel-tabset group="language"}

## Python
```{python presource}

package_name = "black-bear-hunting-activity-and-harvests"

package_response = requests.get(
    f"{api_base}/package_show",
    params={"id": package_name},
    timeout=30
)
package_response.raise_for_status()

package = package_response.json()["result"]
resources = pd.DataFrame(package["resources"])

resource_summary = resources[
    [
        "id",
        "name",
        "format",
        "language",
        "datastore_active",
        "data_last_updated"
    ]
]

resource_summary

```

## R
```{r rresource, message = FALSE, warning = FALSE}

package_name <- "black-bear-hunting-activity-and-harvests"

package_url <- paste0(
  api_base,
  "/package_show?id=", URLencode(package_name, reserved = TRUE)
)

package <- fromJSON(package_url, simplifyDataFrame = TRUE)$result
resources <- as_tibble(package$resources)

resource_summary <- resources |>
  select(
    id,
    name,
    format,
    language,
    datastore_active,
    data_last_updated
  )

resource_summary

```
:::

For DataStore API calls, choose a resource where `datastore_active` is `TRUE`. This dataset has English and French CSV resources. The English CSV resource id is the value used in the data download.

::: {.panel-tabset group="language"}

## Python
```{python pselect}

datastore_resources = resources[
    (resources["datastore_active"] == True) &
    (resources["format"] == "CSV")
]

english_resource = datastore_resources[
    datastore_resources["language"] == "english"
].iloc[0]

resource_id = english_resource["id"]

resource_id

```

## R
```{r rselect, message = FALSE, warning = FALSE}

datastore_resources <- resources |>
  filter(datastore_active == TRUE, format == "CSV")

english_resource <- datastore_resources |>
  filter(language == "english") |>
  slice(1)

resource_id <- english_resource$id

resource_id

```
:::

### Organizing Dataset

The following code is used to download the dataset. Rather than reading a static CSV snapshot, we use Ontario's CKAN DataStore API. The resource id discovered above identifies the black bear hunting activity table, and the pagination loop retrieves all available records.

::: {.panel-tabset group="language"}

## Python
```{python pdataset}

datastore_search_url = f"{api_base}/datastore_search"

def get_all_records(resource_id, limit=1000):
    records = []
    offset = 0

    while True:
        response = requests.get(
            datastore_search_url,
            params={
                "resource_id": resource_id,
                "limit": limit,
                "offset": offset
            },
            timeout=30
        )
        response.raise_for_status()

        page = response.json()["result"]["records"]
        records.extend(page)

        if len(page) < limit:
            break

        offset += limit

    return records

data_download = pd.DataFrame(get_all_records(resource_id))

total = data_download[data_download.WMU == "Total"]
data = data_download[data_download.WMU != "Total"]

```

## R
```{r rdataset, message = FALSE, warning = FALSE}
datastore_search_url <- paste0(api_base, "/datastore_search")

get_all_records <- function(resource_id, limit = 1000) {
  records <- list()
  offset <- 0

  repeat {
    page_url <- paste0(
      datastore_search_url,
      "?resource_id=", resource_id,
      "&limit=", limit,
      "&offset=", offset
    )

    page <- fromJSON(page_url, simplifyDataFrame = TRUE)$result$records
    records[[length(records) + 1]] <- page

    if (nrow(page) < limit) {
      break
    }

    offset <- offset + limit
  }

  bind_rows(records)
}

data_download <- get_all_records(resource_id)

total <- data_download |> rename_all(make.names)|> filter(WMU =="Total")
data <- data_download |> rename_all(make.names)|> filter(WMU !="Total")

```
:::

The data has columns WMU, year, number of active hunters, and the harvest.  All numbers of bear harvested and active hunters are estimated based on the replies received from a sample of hunters, so it might contain a statistical error.  In the WMU called Total, all of the WMUs are summed.

::: {.panel-tabset group="language"}

## Python
```{python}
# view the first rows
total.head()
# viewthe last rows
data.tail()
```

## R
```{r}
# view the first rows
head(total)
# view the last rows
tail(data)
```

:::

### Plotting the Number of Bear Hunters and Harvest

The following code plots the number of bears harvested per active hunter in each WMU.

::: {.panel-tabset group="language"}

## Python
```{python pplots}

harvest_hunter_plot = sb.scatterplot(data = data,
                                     x = 'Active Hunters',
                                     y = 'Harvest', 
                                     hue = 'WMU', 
                                     legend = False)
harvest_hunter_plot.set_title("Hunters and Harvest in different WMUs");

```

## R
```{r rplotss}

ggplot(data, aes(x = Active.Hunters, y = Harvest, colour = WMU))+
  geom_point(show.legend = FALSE)+
  ggtitle("Hunters and Harvest in different WMUs")
  
```
:::
