---
title: "COVID Data: COVID-19 Data - Government of Canada and JHU CSSE"
author: "Junpu Xie & Dave Campbell"
date: '2020-07-24'
slug: covid-data-covid-19-govenrmne-of-canada-and-jhu-csse
categories: ["COVID Data"]
tags: ["Linear Regression", "Normal Distribution","Time Series", "Aggregated Ddata", "Spatial Data Set", "COVID-19", "COVID Data"]
subtitle: ''
summary: 'JHU CSSE and JHU APL have cooperated together to create a large COVID-19 data repository and update real-time official COVID-19 data worldwide. Moreover, the COVID-19 data from the Government of Canada provide detailed information about COVID-19 data in time-series and aggregated forms.'
lastmod: '2020-07-24T18:48:09-04:00'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Possible Analysis and Keywords
- Linear Regression
- Normal Distribution
- Time Series
- Aggregated Data
- Spatial Data Set
- COVID-19
- COVID Data

## Data Provider - Worldwide 

COVID-19 data plays a vital role in managing a pandemic. Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE), ESRI Living Atlas Team and the Johns Hopkins University Applied Physics Lab (JHU APL) have compiled a COVID-19 data repository which is updated in real-time with worldwide COVID-19 [data](https://github.com/CSSEGISandData/COVID-19).

The real-time visual dashboard can be viewed in [desktop](https://www.arcgis.com/apps/opsdashboard/index.html#/bda7594740fd40299423467b48e9ecf6) and [mobile](http://www.arcgis.com/apps/opsdashboard/index.html#/85320e2ea5424dfaaa75ae62e5c06e61).

## Data Provider - Canada

COVID-19 is a serious health threat for individuals and its extended social problems are also evolving daily. Based on the given number of cases in Canada, the risk to Canadians is still considered high. 

The government of Canada updates regular COVID-19 information and news [here](https://www.canada.ca/en/public-health/services/diseases/coronavirus-disease-covid-19.html), it also creates interactive data visualizations by providing a visual data gallery [here](https://health-infobase.canada.ca/covid-19/visual-data-gallery/), so that everyone can easily see the current situation of COVID-19 across Canada and the world with different measurements. Health Canada provides several visualizations including the [Canada COVID-19 Situational Awareness Dashboard](https://experience.arcgis.com/experience/2f1a13ca0b29422f9b34660f0b705043/?fbclid=IwAR0k_pf1jxwanqldeaZy0-1bao8njVXrz6A3-ejp72Kyh4Qcj0MrE72lgdo) and [interactive data map](https://www.canada.ca/en/public-health/services/diseases/2019-novel-coronavirus-infection.html).

***

## COVID-19 Data from the Government of Canada

The COVID-19 data from the Government of Canada provide detailed information in number, percent, and rate data of tested, confirmed, recovered, and death cases in time series, it is also aggregated forms with a given date among different provinces and you can view and download the dataset [here](https://health-infobase.canada.ca/src/data/covidLive/covid19.csv).

### Organizing Data

The following code is used to download and rearrange the original dataset from webpage.

```{r ontario}

library(reshape2)
library(gridExtra)
library(tidyverse)
library(lubridate)
library(ggpubr)

Canadian.data <- read_csv("https://health-infobase.canada.ca/src/data/covidLive/covid19.csv") %>% 
                 mutate(date = as_date(date, format = "%d-%m-%y"))
```

> Summary

*  We use mutate to modify a column.  Its arguments are of the form new.column.name=function.of.old.column.names.

* the lubridate package includes a lot of functionality for dealing with dates.  Here we use it to convert a character column into date format.


### Different Characteristics of COVID-19 Cases

Consider the number of case types in a province over the past few months

```{r stacked barplot, cache = TRUE, warning = FALSE}
location ="Ontario"
Canadian.data %>% subset(prname == location) %>%
  select(date,numrecoveredtoday,numdeathstoday,numactive)%>%
  rename(recovered.today=numrecoveredtoday,deaths.today=numdeathstoday,active=numactive)%>%
  gather(key = casetype, value = "cases",-date) %>%
ggplot( ) +
  geom_line(aes(x = date, y=cases,colour=casetype)) +
  scale_x_date(date_labels = "%d/%m", date_breaks = "2 weeks") +
  ggtitle(paste0("Daily Cases in ",location))
  


Canadian.data %>% subset(prname == location) %>%
  select(date,
         numtotal_last14, numdeaths_last14)%>%
  gather(key = casetype, value = "cases",-date) %>%
ggplot( ) +
  geom_line(aes(x = date, y=cases,colour=casetype)) +
  scale_x_date(date_labels = "%d/%m", date_breaks = "2 weeks") +
  ggtitle(paste0("Daily Cases in ",location, " 14 day totals"))
 



Canadian.data %>% subset(prname == location) %>%
  select(date,numconf,numdeaths,numrecover)%>%
  mutate(Recovered.plus.Deaths = numdeaths+numrecover)%>%
  gather(key = casetype, value = "cases",-date) %>%
ggplot( ) +
  geom_line(aes(x = date, y=cases,colour=casetype)) +
  scale_x_date(date_labels = "%d/%m", date_breaks = "2 weeks") +
  ggtitle(paste0("Cumulative cases in ",location))



Canadian.data %>% subset(prname == location) %>%
  select(date,numtestedtoday,numtoday)%>%
    gather(key = casetype, value = "cases",-date) %>%
ggplot( ) +
  geom_line(aes(x = date, y=cases,colour=casetype)) +
  scale_x_date(date_labels = "%d/%m", date_breaks = "2 weeks") +
  ggtitle(paste0("Number of tests and new cases per day in ",location))


```

### Make each plot its own facet.  
Look closely at the vertical scales in these plots.



```{r}
  
Canadian.data %>% subset(prname == location) %>%
  select(date,numtestedtoday,numtoday,numdeathstoday,numrecoveredtoday) %>%
  gather(key = casetype, value = "cases",-date) %>%
ggplot( ) +
  geom_line(aes(x = date, y=cases,colour=casetype)) +
  scale_x_date(date_labels = "%d/%m", date_breaks = "2 weeks") +
  ggtitle(paste0("Number of tests and new cases per day in ",location))+
    facet_grid(rows = vars(casetype))

```




### Adjusting the scales.  
Alternatively we can make each figure with its own scale and ggarrange them together.


```{r}
data2use = Canadian.data %>% subset(prname == location) %>%
  select(date,numtestedtoday,numtoday,numdeathstoday,numrecoveredtoday)
#  gather(key = casetype, value = "cases",-date) 

 
p1 = data2use %>%
ggplot( ) +
  geom_line(aes(x = date, y=numtestedtoday),colour="red") + 
  scale_x_date(date_labels = "%d/%m", date_breaks = "4 weeks") +
  ggtitle(paste0("Number of tests in ",location))



p2 = data2use %>%
ggplot( ) +
  geom_line(aes(x = date, y=numtoday),colour="blue") + 
  scale_x_date(date_labels = "%d/%m", date_breaks = "4 weeks") +
  ggtitle(paste0("Number of new cases in ",location))

  
p3 = data2use %>%
ggplot( ) +
  geom_line(aes(x = date, y=numdeathstoday),colour="green") + 
  scale_x_date(date_labels = "%d/%m", date_breaks = "4 weeks") +
  ggtitle(paste0("Number of deaths day in ",location))

  
  
p4 = data2use %>%
ggplot( ) +
  geom_line(aes(x = date, y=numrecoveredtoday),colour="black") + 
  scale_x_date(date_labels = "%d/%m", date_breaks = "4 weeks") +
  ggtitle(paste0("Number recovered day in ",location))

  
  
ggarrange(p1, p2, p3, p4,
                    ncol = 2, nrow = 2)
```


## We can also transform the vertical scale
```{r}

Canadian.data %>% subset(prname == location) %>%
  select(date,numtestedtoday,numtoday,numdeathstoday,numrecoveredtoday) %>%
ggplot( ) +
  geom_line(aes(x = date, y=numdeathstoday),colour="red") + 
  scale_y_log10()+
  scale_x_date(date_labels = "%d/%m", date_breaks = "4 weeks") +
  ggtitle(paste0("Log Number of Deaths in ",location))


```


***


## Ontario School Covid reports:

Ontario keeps [records](https://data.ontario.ca/dataset/summary-of-cases-in-schools) of the number of cases in different schools.  Here we will directly download the latest data from today

```{r}

school = read_csv("https://data.ontario.ca/dataset/b1fef838-8784-4338-8ef9-ae7cfd405b41/resource/8b6d22e2-7065-4b0f-966f-02640be366f2/download/schoolsactivecovid.csv")


school %>% glimpse

```

# daily totals

There are a heck of a lot of school boards and they have long names that will make legends really busy.  Let's start by deleting the words "District", "School" "Board", "scolaire" "of" and a handful of other stop words that could readily be inferred.


```{r}
#ensure all variable names are lower case
colnames(school) = tolower(colnames(school))
school = school %>% mutate(school_board = str_replace_all(
  school_board, pattern = "(\\sof\\s)|([Dd]istrict)|(School)|(Board)|([cC]onseil)|(scolaire)|(\\\xe9coles)|( du )|(conseil)|(\\sde\\s)|(\\sdes\\s)|(\\<district\\>)|
  (\\\x92)",
  replacement = ""))  

```

#Plot within district
Now we can plot the school district data

```{r}

school%>% group_by(reported_date,school_board) %>%
  summarise(dailytotal = sum(total_confirmed_cases)) %>%
  ggplot(aes(x= reported_date, y =dailytotal , colour = school_board)) +
  geom_line() +
  scale_x_date(date_labels = "%d/%m", date_breaks = "1 weeks")+
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

## COVID-19 Packages in R 

There exist a package of time-series worldwide data about reported cases for COVID-19 from the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE) [data repository](https://github.com/CSSEGISandData/COVID-19) in R Studio.


A variety of COVID-19 packages can be found by clicking [Packages] section in the lower right corner panel of the default R studio interface and [Install] in the upper left corner of the Packages section. When you type 'COVID-19' in the package section, you can find several COVID-19 packages. The one used below is [covid19.analytics] from  JHU CSSE. 

The package of [covid19.analytics] contains many simple and useful functions that can help users easily get a powerful visual gallery. The following code shows some simple visualization and modeling functions, more functions of this package can be discovered [here](https://www.rdocumentation.org/packages/covid19.analytics/versions/1.0).

```{R package, cache = TRUE, message = FALSE, warning = FALSE}
library(covid19.analytics)

```



# Pulling Data
The [covid19.data] function pulls the latest data from JHU CSSE. 
Data options include [aggregated] for all cases aggregated by country or a variety of time series for confirmed cases, deaths, etc...  The package vignette also allows users to obtain covid19's genomic sequence data from NCBI, but we won't do that here.

```{r datagrab, cache=TRUE, eval=FALSE}
aggretated   <- covid19.data(case = "aggregated") %>% as_tibble()
confirmed_US <- covid19.data(case = "ts-confirmed-US") %>% as_tibble()
deaths_US    <- covid19.data(case = "ts-deaths-US") %>% as_tibble()

```


## Auto reports

The package will perform a variety of standard reports and provide some standard models. Here is a basic summary report for North America.  This function will run the report but it is a lot of info for this little html page.

```{r reports, cache=TRUE, eval = FALSE}

report.summary(geo.loc="NorthAmerica", graphical.output = TRUE)
#report.summary(geo.loc="Canada", graphical.output = TRUE)
#report.summary(geo.loc="US", graphical.output = TRUE)


```

