---
title: "Annual and Monthly Soft Drinks Production - Open Canada"
author: "Junpu Xie & Dave Campbell"
date: '2020-08-10'
slug: manufacture-data-annual-monthly-soft-drinks-production-open-canada
categories: ["Regression"]
tags: ["Linear Chart", "Data Merging and Cleaning", "Time Series", "Soft Drinks", "Canada"]
subtitle: ''
summary: 'The dataset of Soft drinks production consists of dates and production variables, and they are recorded by the unit of thousand gallons on a monthly, quarterly, and annual basis.'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Possible Analysis and Keywords

- Linear Chart
- Time Series
- Data Merging and Cleaning
- Soft Drinks
- Manufacturing Data
## Data Provider

Statistics Canada's [Open Government](https://open.canada.ca/en) is a free and open-access platform  containing over 80,000 datasets across diverse subjects. The purpose of sharing all data documents with the public is to remain transparent and accessible. 

Dataset can be discovered by multiple searching methods [here](https://open.canada.ca/en/open-data), such as [Browse by subject](https://open.canada.ca/en/open-data), [Open Government Portal](https://search.open.canada.ca/en/od/?sort=last_modified_tdt%20desc&page=1&search_text=&od-search-portal=Open%20Data&_ga=2.198521263.1872662645.1597195650-535176367.1594143568) for direct keywords search, [Open Maps](https://open.canada.ca/en/open-map) which contains geospatial information data, [Open Data Inventory](https://open.canada.ca/en/search/inventory) from the government of Canada organization, [Apps Gallery](https://open.canada.ca/en/apps) for representing those mobile and web-based application data, [Open Data 101](https://open.canada.ca/en/open-data-principles) for letting people know how to use dataset and so on.

***

## Food Manufacturing: Annual Soft Drink Production 

A dataset recording the soft drink production every quarter from 1950 to 1977, its supporting documentation can be found [here](https://open.canada.ca/data/en/dataset/15dac000-240f-494c-86b6-311d8f8012b9), you also can quickly discover the customized table [here](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1610010001).

The dataset switched to monthly soft drinks production from 1976 to 1995. Its supporting documentation can be found [here](https://open.canada.ca/data/en/dataset/de2e4ec3-dcb8-4dd1-aad8-3ccc1a61d633), you also can quickly discover the customized table [here](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1610009901).

All datasets and their metadata files including detailed variable descriptions are stored as a zip file. The variables in the data sets are dates and production variables recorded in units of thousands of gallons on a monthly, quarterly, and annual basis.

## Exprolary Analysis

### Data loading 

The following code is used to organize monthly and annual orginal datasets, and merge them into one quarterly soft drink production dataset from 1950 to 1995. 

```{r dataset, message = FALSE}

# library
library(tidyverse)
library(lubridate)

#Download the zip file of QUARTERLY soft drink production AND rearrange the dataset
temp <- tempfile()
download.file("https://www150.statcan.gc.ca/n1/tbl/csv/16100100-eng.zip",temp)
(file_list <- as.character(unzip(temp, list = TRUE)$Name))
quarterlydata <- read_csv(unz(temp, "16100100.csv"))
unlink(temp) # Delete temp file

# Organize the quarterly dataset 
quarterly     = quarterlydata 
quarterly[,1] = as_date(paste(quarterly[,1],"-1",sep =""))
quarterly     = quarterly %>% rename(quarter = REF_DATE, QuarterlyValue = VALUE)%>%
                select(quarter, QuarterlyValue)



#Download the zip file of MONTHLY soft drink production AND rearrange the dataset
temp <- tempfile()
download.file("https://www150.statcan.gc.ca/n1/tbl/csv/16100099-eng.zip",temp)
(file_list <- as.character(unzip(temp, list = TRUE)$Name))
monthlydata <- read_csv(unz(temp, "16100099.csv"))
unlink(temp)
monthlydata     = monthlydata %>% rename(date = REF_DATE, Produced = VALUE)


# It is necessary to add the month and day values at the end of the year data before running as.Date() function. Since the original CSV file has its default data format, we need to fill out the missing parts of the date, which is date data in this case, as the character format, then use the as.Date() function to convert. 

monthly <- monthlydata
                mutate(date =  as_date(paste(monthlydata[,"date"],"-1",sep ="")))%>%
                select(date, Produced) 
                

# Now we want to extend the quarterly data set by combining over the monthly dataset:             


MoreQuarters <- monthly %>% 
                mutate(quarter =  quarter(date,with_year = TRUE)) %>%
                group_by(quarter) %>%
                mutate(QuarterlyValue = sum(Produced))%>%
                select(date, quarter, QuarterlyValue) %>%
                distinct( quarter, QuarterlyValue, .keep_all=TRUE) %>%
                ungroup() %>%
                select(date, QuarterlyValue) %>%
                rename(quarter = date)
                
AllQuarters = rbind(quarterly,MoreQuarters)



# Create the dataset of annual soft drink production by summarizing the quaterly production value by year
annual = AllQuarters %>% 
  mutate(Year = floor_date(AllQuarters$quarter, unit = "year")) %>%
  group_by(Year) %>%
  mutate(AnnualValue = sum(QuarterlyValue))%>%
  select(Year, AnnualValue) %>%
  distinct( Year, AnnualValue) 


head(AllQuarters)
head(annual)
head(monthly)
```


## Soft drink production

```{r linear}

### Plot the graph of annual and quaterly soft drink production 
ggplot(annual, aes(x = Year, y = AnnualValue, group = 1)) + 
  geom_line() + 
  ggtitle("Annual soft drink production")


ggplot(AllQuarters, aes(x = quarter, y = QuarterlyValue, group = 1)) + 
  geom_line() +
  geom_point(position="jitter") + 
  ggtitle("Quaterly soft drink production")


ggplot(monthly, aes(x = date, y = Produced, group = 1)) + 
  geom_line() + 
  ggtitle("Monthly soft drink production")

```

* What happened to the annual data from 1976 and 1977?  Need a hint, Zoomm in by changing the x axis limits by adding this to the ggplot:
```{r}
#xlim(as_date(c("1970-01-01","1980-12-01")))
```
* Need another hint?  Look back at the quarterly data sets before they were combined from different sources.


## Quarterly soft drink production, just the old dataset

To better see the production proportions across the four quarters in each year, the following code is used to plot the stacked percentage bar plots of quarterly soft drink production.

* Check out how the Year-Month-Date is split into its elements using _separate_:

```{r barplot}

# Use separate() function form tidyr package to take the date variable apart into year and month

 quarterly %>% 
  separate( quarter, into = c("year","month", "date"), sep = "-") %>%
# Stacked + percent
ggplot( aes(fill=month, y= QuarterlyValue, x=as.Date(year,"%Y"))) +
  geom_bar(position="fill", stat="identity") + 
  ggtitle("Percent Stacked Barchart of Quaterly Soft Drink Production") +
  xlab("Year") + 
  ylab("Percentage")+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Summary

* The percent produced in each quarter is reasonably stable. 
