---
title: Open Ottawa
author: "Dean Jayatilleke & Junpu Xie & Dave Campbell"
date: '2024-07-10'
slug: 
categories: []
tags: ["Ottawa"]
subtitle: ''
summary: 'Retrieving Open Ottawa Data and producing plots using Open Street Maps'
output: html_document
ipynb-shell-interactivity: all
---

## Possible Analysis and Keywords

- Open Street Maps
- Traffic Collisions
- Service Requests
- Ottawa Open Data

## Data Provider

[Open Ottawa](https://open.ottawa.ca/) is an open-access data portal designed to provide accessible datasets from municipal data sources. Users can access datasets from a variety of municipal departments including Parks, Public Health, Water Treatment, and many others. All data from this portal is subject to [Open Data Licence Version 2.0](https://open.ottawa.ca/pages/open-data-licence). Information released under this licence may be copied, modified, published, adapted, or otherwise used for any lawful purpose, as long as a data attibution from the information provider is included. When using information from multiple providers, the following attribution must be used: Contains information licensed under the Open Government Licence – City of Ottawa.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

## Plotting Geographical Data Points on Open Street Maps

```{r libraries, message = FALSE, warning = FALSE}

library(leaflet) 
library(sf)
library(tidyverse)

```

A word about retrieving Open Ottawa Data.

Open Ottawa datasets are available to download as CSVs, however the download URL is generated on demand, and expires shortly, within a few days,
or sometimes even hours. For this reason, when revisiting a dataset, you may have to navigate to the dataset's "About" page and regenerate a download link.

### Service Requests 

```{r service request dataset}

#https://open.ottawa.ca/datasets/8a5030af268a4a3485b72356dd7dfa85/about

csv_path <- ("311opendata_currentyear")

service <- read_csv(csv_path)

head(service)

```

```{r service data prep}

colnames(service) <- sub(" \\|.*", "", colnames(service))
service <- service |> rename_all(make.names)

```

Some service requests do not include location data. Let's filter those out.

```{r service locations}

service_locations <- service |> filter(!str_detect(Latitude, 'N'))

```

After initially plotting all points, I noticed that there was a service request apparently placed from the Pacific Ocean near Africa. Clearly there was a data input error. It's worth filtering for latitudes containing 45 (which reflect Ottawa). We will also filter for a particular type of service request to reduce the total number of points plotted.

```{r park service map}

parkservice <- service_locations |> filter(str_detect(Latitude, "45")) |>
               filter(str_detect(Description, "Park")) |>
               st_as_sf(coords = c("Longitude", "Latitude")) |>
               st_coordinates()

leaflet() |>
  addTiles() |>
  addMarkers(data = parkservice)

```

### Traffic Collisions

```{r traffic collision dataset}

#https://open.ottawa.ca/datasets/ottawa::2020-tabular-transportation-collision-data/about

csv_path <- "2020_Tabular_Transportation_Collision_Data.csv"

traffic = read_csv(csv_path)

head(traffic)

```

Let's filter for a particular type of collison and plot them on Open Street Maps.

```{r sideswipe map}

sideswipe_latlong <- traffic |> filter(str_detect(Initial_Impact_Type, "Sideswipe")) |>
                     st_as_sf(coords = c("Longitude", "Latitude")) |>
                     st_coordinates()
                     
leaflet() |>
  addTiles() |>
  addMarkers(data = sideswipe_latlong)
```