---
title: Ottawa Police Open Data"
author: "Dean Jayatilleke & Junpu Xie & Dave Campbell"
date: '2024-07-10'
slug: 
categories: []
tags: ["Ottawa"]
subtitle: ''
summary: 'Retrieving Ottawa Police Open Data and producing plots using Open Street Maps'
output: html_document
ipynb-shell-interactivity: all
---

## Possible Analysis and Keywords

- Open Street Maps
- Bike Thefts
- Shootings
- Crime
- Motor Vehicle Thefts
- Police Data

## Data Provider

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries}

library(leaflet)
library(sf)
library(tidyverse)

```

```{r shootings dataset}

#https://data.ottawapolice.ca/datasets/aeb09d77912246139014516c8eeefcd9_0/about

csv_link <- "https://stg-arcgisazurecdataprod7.az.arcgis.com/exportfiles-858-723/Shootings_Open_Data_3086844035331583488.csv?sv=2018-03-28&sr=b&sig=T%2FUcCA7fhMR46RZK9zLsnzCFAOqTXeFfUpFOiY8CbkI%3D&se=2024-07-18T18%3A59%3A53Z&sp=r"

shootings <- read_csv(csv_link)

head(shootings)

```

The location of each reported shooting is given in the EPSG:2951 coordinate system, represented in the X and Y columns. 
In order to plot these points on Open Street Maps using the Leaflet library, we will need to convert to Latitude and Longitude, also
known as the EPSG:4386 coordinate system. R's Simple Features (sf) library can be used to manipulate spatial data.
[Simple Features] (https://cran.r-project.org/web/packages/sf/index.html).


