---
title: "Ontario Research Fund - Research Excellence Program - Open Ontario"
author: "Dean Jayatilleke, Junpu Xie, and Dave Campbell"
date: '2024-08-18'
execute:
  render: auto
  slug: education-data-ontario-research-fund-research-excellence-ptrogram-data-catalogue-frome-governmnet-of-ontario
categories: ["Pairs Plot","Open Ontario", "Software:Python","Software:R","zipfiles", "xlsx files", "CKAN API"]
subtitle: ''
summary: 'This dataset includes hundreds of research porjects funded through Research Excellence Program from 2003 to 2018.'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Provider

The provincial government of Ontaio provides open access to thousands of data sets via their [Open Data Ontario](https://data.ontario.ca/en/) portal. The purpose of sharing all data documents with the public is to remain transparent and accessible.  More details about the data license, training materials, and other information can be found [here](https://data.ontario.ca/about).

***

## Ontario Research Funding


This data set provides a summary of research projects funded by the Ministry of Colleges and Universities.  The data format changes in 2018-2019, so the dataset is split into current and historical versions.  We will extract both of them, since they differ in format and API details.  

The Ontario Research Fund [current](https://data.ontario.ca/dataset/ontario-research-funding-summary-current) dataset includes research projects funded through the Research Excellence Program from 2018 to 2026. The Ontario Research Fund [historical](https://data.ontario.ca/en/dataset/ontario-research-funding-summary-historical) dataset  covers 2004 - 2018.  These dataset can be found online as an **xlsx** file so you could use a web browser to visit the page, find the appropriate dataset(s), download, and then load it into R / Python directly.  However, for reproducibility and to facilitate reporting updates when new data is released, the full process is automated using the API.

## Goal

We search for the datasets using the CKAN API.   From there we filter down to the English versions and download the datasets.   Data is in different formats, so we need to extract the appropriate sheet from the xlsx files and merge them to extend the dataset.  Some scatter plots are made.







## Libraries

::: {.panel-tabset group="language"}

## Python
```{python plib}

import pandas as pd   
import seaborn as sb
import requests
import openpyxl
import matplotlib.pyplot as plt

```
## R
```{r rlib, warning = FALSE, message = FALSE}
library(jsonlite)   # handling the json returns from API calls
library(tidyverse)  # data manipulation
library(lubridate)  # dates
library(viridis)    # colours
library(hrbrthemes) # plot options
library(ggridges)   # plot options
library(readxl)     # loading excel files
```
:::


### Using the API to Find the DataStore Resource Id



The Ontario Data Catalogue uses the [CKAN API](https://data.ontario.ca/about#developers). The CKAN API uses some specific terminology, a CKAN package is a dataset landing page, while a resource is an individual file, table, API-backed data object, or supporting document attached to that dataset.  
The CKAN API is similar ot a REST API in that we interface it using a constructed URL and obtain a response.   We will use it to search for a dataset and obtain the url to download a data file.

We use the API to search for the research funding dataset and identify the file url.  Here we'll search for **ontario-research-funding** and see if we can find the data table identifier.   Search is performed by adding search terms to the `api_base` to construct a url.  We include a `timeout` in the request to avoid waiting forever if our internet (or the API) breaks.  The API returns a json file `search_results` with up to 10 rows (user defined value that works here because there aren't many bear datasets) that needs reshaping to make it easier to read the important values.

We already know our target dataset is  [https://data.ontario.ca/en/dataset/ontario-research-funding-summary-current](https://data.ontario.ca/en/dataset/ontario-research-funding-summary-current), so the package name is `ontario-research-funding-summary-current`. But to showcase the process, we will start by searching available resources to find the dataset name. If we knew the package name (from navigating in a browser) we could query the API via `package_show?id=<exact resource name here>`.   Here we assume that we don't know the name and need to search for it using the API to query via `package_search?q=<keywords go here>`.
  



::: {.panel-tabset group="language"}

## Python
```{python presearch, message = FALSE, warning = FALSE}
api_base = "https://data.ontario.ca/en/api/3/action"
research_search_terms = "ontario-research-funding"

research_search_response = requests.get(
    f"{api_base}/package_search",
    params={
        "q": research_search_terms,
        "rows": 200
    },
    timeout=30
)
research_search_response.raise_for_status()

research_search_results = research_search_response.json()["result"]["results"]

research_datasets = pd.DataFrame([
    {
        "name": dataset["name"],
        "title": dataset["title"],
        "resources": dataset["num_resources"],
        "modified": dataset["metadata_modified"]
    }
    for dataset in research_search_results
])

research_matches = research_datasets[
    research_datasets["name"].str.contains(
        "ontario-research-funding",
        regex=False
    )
]

research_matches


```

## R
```{r rresearch, message = FALSE, warning = FALSE}
api_base = "https://data.ontario.ca/en/api/3/action"
research_search_terms <- "ontario-research-funding"

research_search_url <- paste0(
  api_base,
  "/package_search?q=", URLencode(research_search_terms, reserved = TRUE),
  "&rows=200"
)

research_search_results <- fromJSON(
  research_search_url,
  simplifyDataFrame = TRUE
)$result$results

research_search_results|> glimpse()

research_datasets <- research_search_results |>
  transmute(
    name,
    title,
    resources = num_resources,
    modified = metadata_modified
  )

research_matches <- research_datasets |>
  filter(str_detect(name, fixed("ontario-research-funding")))

research_matches

```
:::

We've managed to find the two different datasets; **current** and **historical**.  We will start by selecting the **current** using a regular expression match, and pass the data package name into another API call in the field `package_show`.  Note that  `resources = 2`, which after filtering to **current** below, we will see shortly relates to English and French versions of the dataset.  Later we will come back to this step and extract the **historical** dataset through a related but somewhat different procedure.  


### Current dataset (obtain and extract the xlsx file) 



::: {.panel-tabset group="language"}

## Python
```{python presearch2, message = FALSE, warning = FALSE}
research_package_name = research_matches[
    research_matches["name"].str.contains("current", regex=False)
].iloc[0]["name"]

research_response = requests.get(
    f"{api_base}/package_show",
    params={"id": research_package_name},
    timeout=30
)
research_response.raise_for_status()

research_package = research_response.json()["result"]
research_resources = pd.DataFrame(research_package["resources"])

research_resource_summary = research_resources[
    [
        "id",
        "name",
        "format",
        "language",
        "datastore_active",
        "data_last_updated",
        "url"
    ]
]

research_resource_summary
```

## R
```{r rresearch-search, message = FALSE, warning = FALSE}
research_package_name <- research_matches |>
  filter(str_detect(name, fixed("current"))) |>
  slice(1) |>
  pull(name)

research_package_url <- paste0(
  api_base,
  "/package_show?id=", URLencode(research_package_name, reserved = TRUE)
)

research_package <- fromJSON(research_package_url, simplifyDataFrame = TRUE)$result
research_resources <- as_tibble(research_package$resources)

research_resource_summary <- research_resources |>
  select(
    id,
    name,
    format,
    language,
    datastore_active,
    data_last_updated,
    url
  )

research_resource_summary

```
:::
  
The research funding package has resources, but they are _Excel files_ and `datastore_active` is **FALSE**. That means the `datastore_search` code used for the bear hunting data is not the right next step. In this case, the CKAN API is used to find the current download URL.  

Excel files are like complicated lists of csv files.  A typical workflow will always involves familiarizing yourself with the datafiles.  In this example we will assume that we will recognize the `sheet` name.   I know from exploring the dataset elsewhere that the sheet name changes as the data updates, so we will do some exploring to find the sheet names before settling into making a dataframe of the relevant sheet.  There are a lot of bonus options when reading in from Excel (or a general csv file), such as the number of rows to skip, loading a subset of columns..., so keep that in mind for more complex analysis. 

**Note**, there is a minor difference between how R and python obtain and load the Excel file.  Python will download and open the Excel file directly from the url, whereas in R we need to download the file into a temporary location and load it from there.  


::: {.panel-tabset group="language"}
## Python
```{python presearch-download-url}

# filter down to English
# also note that I'm being lazy and extracting the first row of a 1 row data frame
# and that this is hard coded with iloc[0].  That may not be good for your needs.
research_download_url = research_resources[
    (research_resources["language"] == "english") &
    (research_resources["format"] == "XLSX")
].iloc[0]["url"]

# the url:
research_download_url

# list the sheet names:
with pd.ExcelFile(research_download_url) as xls:
    print(xls.sheet_names)

# load only the sheet name that we recognize.  
# Seeing our sheet_name = 'July 2018-Jan 31 2026', we could use that name explicitly, 
# but since the sheet name will update when new data comes out we just hope that the 
# format doesn't change and ask for the sheet in position 1
research_data_current = pd.read_excel(research_download_url, sheet_name = 1)
research_data_current.head()
```

## R
```{r rresearch-download-url, message = FALSE, warning = FALSE}

research_download_url <- research_resources |>
  filter(language == "english", format == "XLSX") |>
  slice(1) |>
  pull(url)
# the url:
research_download_url

# To load the file in R, you need to download it first, then open it using read_excel
dest_file <- tempfile(fileext = ".xlsx")
download.file(research_download_url, destfile = dest_file, mode = "wb")

# now we can explore the Excel file.  
# list the sheet names:
excel_sheets(dest_file)

# load only the sheet name that we recognize.  
# Seeing our sheet_name = 'July 2018-Jan 31 2026', we could use that name explicitly, 
# but since the sheet name will update when new data comes out we just hope that the 
# format doesn't change and ask for the sheet in position 2

research_data_current <- readxl::read_excel(dest_file, sheet = 2)
research_data_current |> head()
```
:::







### Historical dataset 

We already have the search information from earlier, now we repeat the process changing **current** to **historical** to obtain the second datafile.



::: {.panel-tabset group="language"}

## Python
```{python presearch2-historic, message = FALSE, warning = FALSE}
research_package_name = research_matches[
    research_matches["name"].str.contains("historical", regex=False)
].iloc[0]["name"]

research_response = requests.get(
    f"{api_base}/package_show",
    params={"id": research_package_name},
    timeout=30
)
research_response.raise_for_status()

research_package = research_response.json()["result"]
research_resources = pd.DataFrame(research_package["resources"])

research_resource_summary = research_resources[
    [
        "id",
        "name",
        "format",
        "language",
        "datastore_active",
        "data_last_updated",
        "url"
    ]
]

research_resource_summary
```

## R
```{r rresearch-search-historic, message = FALSE, warning = FALSE}
research_package_name <- research_matches |>
  filter(str_detect(name, fixed("historical"))) |>
  slice(1) |>
  pull(name)

research_package_url <- paste0(
  api_base,
  "/package_show?id=", URLencode(research_package_name, reserved = TRUE)
)

research_package <- fromJSON(research_package_url, simplifyDataFrame = TRUE)$result
research_resources <- as_tibble(research_package$resources)

research_resource_summary <- research_resources |>
  select(
    id,
    name,
    format,
    language,
    datastore_active,
    data_last_updated,
    url
  )

research_resource_summary

```
:::
  
As before, the research funding package has resources, but they are _Excel files_ and `datastore_active` is **FALSE**. We use the CKAN API to find the download URL.  


::: {.panel-tabset group="language"}
## Python
```{python presearch-download-url-historic}

# filter down to English
# also note that I'm being lazy and extracting the first row of a 1 row data frame
# and that this is hard coded with iloc[0].  That may not be good for your needs.
research_download_url = research_resources[
    (research_resources["language"] == "english") &
    (research_resources["format"] == "XLSX")
].iloc[0]["url"]

# the url:
research_download_url

# list the sheet names:
with pd.ExcelFile(research_download_url) as xls:
    print(xls.sheet_names)

# load only the sheet name that we recognize.  
# Seeing our sheet_name = 'Oct 2004-June 2018', 
# since the sheet name will probably not update since the data is historical, 
# we hard code the name 
research_data_historical = pd.read_excel(research_download_url, sheet_name = 'Oct 2004-June 2018')
research_data_historical.head()
```

## R
```{r rresearch-download-url-historic, message = FALSE, warning = FALSE}

research_download_url <- research_resources |>
  filter(language == "english", format == "XLSX") |>
  slice(1) |>
  pull(url)
# the url:
research_download_url

# To load the file in R, you need to download it first, then open it using read_excel
dest_file <- tempfile(fileext = ".xlsx")
download.file(research_download_url, destfile = dest_file, mode = "wb")

# now we can explore the Excel file.  
# list the sheet names:
excel_sheets(dest_file)

# load only the sheet name that we recognize.  
# Seeing our sheet_name = 'Oct 2004-June 2018', 
# since the sheet name will probably not update since the data is historical, 
# we hard code the name 


research_data_historical <- readxl::read_excel(dest_file, sheet = 'Oct 2004-June 2018')
research_data_historical |> head()
```






### Tips based on status of datastore\_active

- If a resource has `datastore_active == TRUE`, use its `id` with `datastore_search`.
- If `datastore_active == FALSE`, use the resource `url` directly to download the file.   Load the datafile appropriately for its `format`, typically from Open Ontario we are interested in CSV, XLSX, GeoJSON, or ZIP.
- If a package has several resources, filter by `language`, `format`,  `name`, etc...  





### Data Merging


Merging is easiest if the datasets have the same columns, since then they can just be stacked.  Let's look at this column names that match and also those that appear in only one.



::: {.panel-tabset group="language"}
## Python
```{python psamesame}
# Columns common to both DataFrames
print(list(
    set(research_data_historical.columns)
    .intersection(research_data_current.columns)
))


# Columns in current but not historical
print(list(
    set(research_data_current.columns)
    .difference(research_data_historical.columns)
))


# Columns in historical but not current
print(list(
    set(research_data_historical.columns)
    .difference(research_data_current.columns)
))


```


## R
```{r rsamesame}
intersect(
  colnames(research_data_historical),
  colnames(research_data_current)
)


setdiff(
  colnames(research_data_current),
  colnames(research_data_historical) 
  )

setdiff(
  colnames(research_data_historical),
  colnames(research_data_current)
  )

```
:::

The only difference is a spacing issue in a variable that is not of current interest.   Reformatting the column names would resolve this.   The other potential problem is the data formats don't always align; in R, the `Field of Research (FOR) - Level 1 Division Code` is a double in **historical** but a character in **current**.  To keep things simple we will first filter the dataset and stack only the columns of interest.


### On to making plots

We will focus on the overall support from the `Program` **Ontario Research Fund - Research Excellence** at an `Institution Type` of **University**, and then keep only the `Total Project Costs`, `Ontario Commitment`, and `Fiscal Year`, and we will plot these.

::: {.panel-tabset group="language"}


###%%%###%%##%%## HERE HERE HERE HERE HERE check python if it works, then maybe convert fiscal year to an integer by trimming the "-\\d{2}" and converting to numeric.   Then use the same same plots.

## Python
```{python pdataset}

research_data = pd.concat(
    [research_data_current, research_data_historical],
    axis=0,
    ignore_index=True
)

# Filter the data
df = (
    research_data
    .query("Program == 'Ontario Research Fund - Research Excellence'")
    .query("`Institution Type` == 'University'")
    [["Total Project Costs", "Ontario Commitment", "Fiscal Year"]]
)

# Create plot
fig, ax = plt.subplots(figsize=(8, 6))

sb.scatterplot(
    data=df,
    x="Total Project Costs",
    y="Ontario Commitment",
    hue="Fiscal Year",
    size="Fiscal Year",
    alpha=0.7,
    ax=ax
)

# Add y = x reference line
xmin = 0
xmax = df["Ontario Commitment"].max()

ax.plot(
    [xmin, xmax],
    [xmin, xmax],
    linestyle="--",
    alpha=0.5
)

ax.set_xlabel("Total Project Costs")
ax.set_ylabel("Ontario Commitment")

plt.tight_layout()
plt.show()





# convert to numeric for filtering, then convert back to a factor as plotting hack
plot_data_2021 = df.copy()
plot_data_2021["Fiscal Year"] = (
    plot_data_2021["Fiscal Year"]
    .str.replace(r"-\d{2}", "", regex=True)
    .astype(float)
)

plot_data_2021 = plot_data_2021.query("`Fiscal Year` > 2020")

plot_data_2021["Fiscal Year"] = plot_data_2021["Fiscal Year"].astype(int).astype(str)
fig, ax = plt.subplots(figsize=(8, 6))

sb.scatterplot(
    data=plot_data_2021,
    x="Total Project Costs",
    y="Ontario Commitment",
    hue="Fiscal Year",
    alpha=0.8,
    ax=ax
)


xmax = min(plot_data_2021["Total Project Costs"].max(), plot_data_2021["Ontario Commitment"].max())

ax.plot([0, xmax], [0, xmax], linestyle="--", alpha=0.4)

ax.set_title("Ontario Research Funding Costs and Commitments since 2021")

plt.tight_layout()

plt.show()
```

## R
```{r rdataset, warning = FALSE, message = FALSE}


research_data_c = research_data_current |> 
    subset(Program == "Ontario Research Fund - Research Excellence") |>
    subset(`Institution Type` == "University") |> 
    select(`Total Project Costs`,`Ontario Commitment`, `Fiscal Year`) 

research_data_h = research_data_historical |> 
    subset(Program == "Ontario Research Fund - Research Excellence") |>
    subset(`Institution Type` == "University") |> 
    select(`Total Project Costs`,`Ontario Commitment`, `Fiscal Year`) 

research_data = bind_rows(research_data_c, research_data_h) 

  
  research_data |>
    ggplot(aes(x = `Total Project Costs`, colour = `Fiscal Year`, y = `Ontario Commitment`)) + 
  geom_abline(slope = 1, intercept = 0, alpha = .5, linetype = "dashed") + 
  geom_point(alpha = .3)+
    ggtitle("Ontario Research Funding Costs and Commitments")


  
   research_data |>
    mutate(`Fiscal Year` = str_replace_all(`Fiscal Year`, pattern = "-\\d{2}", replacement = "") |> as.numeric()) |>  
    dplyr::filter(`Fiscal Year` > 2020 ) |>  # convert to numeric for filtering, then convert back to a factor as plotting hack
    mutate(`Fiscal Year` = as.character(`Fiscal Year`)) |>  
    ggplot(aes(x = `Total Project Costs`, colour = `Fiscal Year`, y = `Ontario Commitment`)) + 
  geom_abline(slope = 1, intercept = 0, alpha = .4, linetype = "dashed") + 
  geom_point(alpha = .8)+
    ggtitle("Ontario Research Funding Costs and Commitments since 2021")
```

:::

If Ontario was fully funding these research projects, then the points would fall on the dashed line.   Considering 2021 and beyond, you can see that Ontario capped the projects to maximum contribution of $2,000,000 towards the total project costs.


