---
title: "Private Elementary and Secondary Schools Revenues - Open Canada"
date: '2020-08-10'
author: "Junpu Xie & Dave Campbell"
slug: private-elementary-and-secondary-schools-revenues-open-canada
categories: ["Regression"]
tags: ["Time Series", "Exploratory Analysis", "Canada", "School Revenue", "Education", "Ontario"]
subtitle: ''
summary: 'This dataset contains 138 observational series and several variables, which include the geography of Canada and the other 13 provinces, 9 direct sources of funds and total funded revenue from 1947 to 2002.'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Possible analysis and Keywords

- Time Series
- Spatial Data Set
- Stacked Area Chart
- School Revenue
- Education Data

## Data Provider

The provincial government of Ontaio provides open access to thousands of data sets via their [Open Data Ontario](https://data.ontario.ca/en/) portal. The purpose of sharing all data documents with the public is to remain transparent and accessible.  More details about the data license, training materials, and other information can be found [here](https://data.ontario.ca/about).

***

## Private Elementary and Secondary Schools Revenues, by Direct Source of Funds

This dataset contains 138 observations across provinces and funding sources from 1947 to 2002.

The dataset and its metadata file which contains detailed variable descriptions have been stored together as a zip file. The data resource and its description can be found [here](https://open.canada.ca/data/en/dataset/5b7fd541-5ae6-4602-a895-29210f0abd78), you also can quickly discover the customized table [here](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3710008401).


## Exploratory Analysis

### Organizing Dataset

The following code is used to download and organize the original dataset, and only select the important variables to form a new dataset.

```{r dataset, message = FALSE}

# library
library(tidyverse)
library(viridis)
library(hrbrthemes)

# Download the zip file of school revenues dataset
temp <- tempfile()
download.file("https://www150.statcan.gc.ca/n1/tbl/csv/37100084-eng.zip",temp)
(file_list <- as.character(unzip(temp, list = TRUE)$Name))
revenue <- read.csv(unz(temp, "37100084.csv"), stringsAsFactors = FALSE)
unlink(temp) # Delete temp file

# Organize the dataset

revenue <- revenue %>% 
  separate(REF_DATE, c("year", "year.end"), "/") %>% 
  mutate(year = as.numeric(year))%>%
  drop_na(VALUE) %>%
  as_tibble()


```

### School Revenue by Different Funded Source in Canada

The following code is used to draw the stacked area chart so that we can directly see the time-series trend and proportion of different funding sources over the years.

```{r graph, warning = FALSE, message = FALSE}

# Subset a reveune dataset which only focus on different funding source in Canada and plot the graph
revenue %>% subset( GEO == "Canada" & Direct.source.of.funds != "Total revenues")%>%
ggplot( aes(x=year, y=VALUE, group=Direct.source.of.funds, colour =Direct.source.of.funds)) + 
    geom_line(alpha=0.6 , size=.5) +
    scale_fill_viridis(discrete = T) +
    theme_ipsum() + 
    labs(title = "School Revenue Funded \n from Different Sources")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

### Total School Revenue in Different Provinces

The following code is used to draw the stacked area chart so that we can directly see the time-series trend and proportion in different provinces over the years.

```{r proince, warning = FALSE, message = FALSE}



# Subset a reveune dataset which only focuses on different funding source in Canada and plot the graph
revenue %>% subset( GEO != "Canada" & Direct.source.of.funds == "Total revenues")%>%
ggplot( aes(x=year, y=VALUE, group=GEO, colour =GEO)) + 
    geom_line(alpha=0.6 , size=.5) +
    scale_fill_viridis(discrete = T) +
    theme_ipsum() +
    ggtitle("Total School Revenue Funded \n in Different Provinces")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```

